Supplement: Autopolyploid establishment through polygenic adaptation

### 2024-06-05 doc/paper/supp.tex
up

\documentclass[11pt,a4paper]{article}
\input{preamble.tex}
\usepackage{geometry}
\usepackage{authblk}
\externaldocument[m-]{main}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thealgorithm}{S\arabic{algorithm}}

\begin{document}
\title{Supplement: Autopolyploid establishment through polygenic adaptation}
\author{Arthur Zwaenepoel}
\date{\vspace{-5ex}}
\maketitle
\tableofcontents

@@

@@

\subsection{Inbreeding in autotetraploids \label{sec:tetinbred}}

In polyploids, the inbreeding coefficient $F_i$ does not suffice to describe
the state of homozygosity in individual $i$.
In tetraploids, for instance, we have five distinct homozygosity states, which
we can symbolically represent as $abcd, aabc, aabb, aaab$ and $aaaa$ (in
general, the number of homozygosity states grows according to the partition
function $(1,2,3,5,7,11,15,22,\dots)$).
Representing the probability of being in these five increasingly homozygous
states as $\delta_1, \dots, \delta_5$, we find that the gametic segregation
variance is reduced by a factor
  $$\phi = \delta_1 + \left(1 - \frac1 6\right) \delta_2 + \left(1 - \frac1
        3\right)\delta_3 + \left(1 - \frac1 2\right) \delta_4$$
which is precisely $1-F_i$, as in diploids (see also \cite{moody1993}).
This shows that we do not need to track the array of homozygosity coefficients
in order to compute the segregation variance in a tetraploid family, but only
require the inbreeding coefficients of the parents.
This is a consequence of the fact that, in tetraploids, gametes are diploid.
For higher ploidy levels, one would need to track higher order identity
coefficients.

@@

>>>>

\documentclass[11pt,a4paper]{article}
\input{preamble.tex}
\usepackage{geometry}
\usepackage{authblk}
\externaldocument[m-]{paper}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thealgorithm}{S\arabic{algorithm}}

\begin{document}
\title{Supplement: Autopolyploid establishment through polygenic adaptation}
\author{Arthur Zwaenepoel}
\date{\vspace{-5ex}}
\maketitle
\tableofcontents
\clearpage

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{/home/arthur_z/dev/InfGenetics/doc/img/finsvsinf-polyploids-validation.pdf}
\caption{
Validation of the autotetraploid and triploid infinitesimal model by a
comparison against the discrete locus model with $L=500$ additive loci, showing
the average phenotypic variance in each generation averaged over 10 replicate
simulations for both models.
We assume the initial phenotypic variance to be one for all simulations, and
all replicates are initialized randomly in accordance with this initial
phenotypic variance.
Allelic effects for the discrete locus model are sampled from a Gaussian with
mean $0$ and variance $1/2L$.
\label{fig:fininf}}
\end{figure}

@@

@@

\subsection{Inbreeding in the mixed-ploidy model}

\subsubsection{Effect of inbreeding on segregation variance in autotetraploids \label{sec:tetinbred}}

In polyploids, the inbreeding coefficient $F_i$ does not suffice to describe
the state of homozygosity in individual $i$.
In tetraploids, for instance, we have five distinct homozygosity states, which
we can symbolically represent as $abcd, aabc, aabb, aaab$ and $aaaa$ (in
general, the number of homozygosity states grows according to the partition
function $(1,2,3,5,7,11,15,22,\dots)$).
Representing the probability of being in these five increasingly homozygous
states as $\delta_1, \dots, \delta_5$, we find that the gametic segregation
variance is reduced by a factor
  $$\phi = \delta_1 + \left(1 - \frac1 6\right) \delta_2 + \left(1 - \frac1
        3\right)\delta_3 + \left(1 - \frac1 2\right) \delta_4$$
which is precisely $1-F_i$, as in diploids (see also \cite{moody1993}).
This shows that we do not need to track the array of homozygosity coefficients
in order to compute the segregation variance in a tetraploid family, but only
require the inbreeding coefficients of the parents.
This is a consequence of the fact that, in tetraploids, gametes are diploid.
Similar considerations apply to triploids if we only model haploid and diploid
gametes.
For higher gametic ploidy levels, one would need to track higher order identity
coefficients in polyploids, which is intractable in general \citep{barton2023}.

\subsubsection{Recursions for inbreeding coefficients in the mixed-ploidy model}

There is little difficulty in extending the recursions for diploids
\citep{barton2017} and autotetraploids (\cref{eq:rectet}) to the mixed-ploidy
case. 
Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
inbreeding coefficients in the mixed-ploidy case becomes
\begin{align}
    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \nonumber \\ 
    F_i &= \frac{1}{3}\left(F_k^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 2, g_l = 1 \nonumber \\ 
    F_i &= \frac{1}{3}\left(F_l^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 1, g_l = 2 \nonumber \\ 
    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
\end{align}
where $F_k^\ast = \alpha_{c_k} + (1-\alpha_{c_k})F_k$, as in the
autotetraploid model.
The recursion for the coancestry coefficients in is given by
\begin{align}
    \Phi_{ii} &= \frac{1}{c_{i}} \left(1 + (c_i-1)F_i\right) \nonumber \\
    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
    \label{eq:coancestry}
\end{align}
where the sums are over individuals in the parental population, and where
$P_{ik} \in \{0, \frac1 3, \frac1 2, \frac2 3, 1\}$ is the probability that a
gene copy in $i$ is derived from parent $k$.

Consider a single locus in a population of $k$-ploids, and let $X_1,\dots,X_k$
denote a random genotype at this locus, where $X_i$ is the additive genetic
value of the allele on homolog $i$.
The variance in gametic values $Y$ produced by some meiotic process can be
decomposed as 
\begin{equation}
  \Var[Y] = \underbrace{\Ex\left[\Var[Y|X_1, \dots, X_k]\right]}_{\text{segregation variance}} 
    + \underbrace{\Var\left[\Ex[Y|X_1, \dots, X_k]\right]}_{\text{population
variation}}
\label{eq:eve}
\end{equation}
We use this relationship to derive the segregation variance for a given meiotic
process.

As an example, consider ordinary diploid meiosis at a single locus, where a
diploid produces a haploid gamete. $Y$ is a random haploid gamete sampled from the
population, and $\Var[Y] = \Var[X] = v_x$, where $X$ is a a randomly sampled
gene from the population and $v_x$ is the variance in additive genetic values
at the locus.
We have
  $$\Var[\Ex[Y|X_1,X_2]] = \Var\left[\frac{X_1 + X_2}{2}\right] = \frac{v_x}{2}$$
Using \cref{eq:eve}, we can then find the segregation variance
  $$\Ex[\Var[Y|X_1,X_2]] = v_x - \frac{v_x}{2} = \frac{v_x}{2}$$
Summing over $L$ independent loci, we find the gametic segregation variance in
diploids (which is halve the zygotic segregation variance $V$, by definition) as
  $$V_{S(2,1)} = \frac{V}{2} = \sum_{i=1}^L \frac{v_{x,i}}{2} = \frac{V_{x,2}}{2}$$
and hence $V=V_{x,2}$, where $V_{x,2}$ is as defined in \cref{sec:scaling}.
This will hold in the infinitsimal limit, where $L$ becomes very large and
$v_x$ smaller and smaller.

Below, we derive the segregation variance associated with haploid and diploid
gamete production in the three cytotypes of the mixed-ploidy population.
We use \cref{eq:eve} to do this, and express the segregation variance for a
$k$-ploid individual producing an $l$-ploid $V_{S(k,l)}$ in terms of the
variance $V_{x,k}$ associated with a haploid genome in a (hypothetical)
$k$-ploid reference population at HWLE. Using the scaling assumptions above,
i.e. $V_{x,k} = \beta_k^2 V$, we can then express all segregation variances in
terms of the diploid segregation variance in the reference population $V$.

\paragraph{Meiosis in autotetraploids.}

When an autotetraploid forms quadrivalents during prophase I, a form of
'internal inbreeding' may occur as a result of the phenomenon called
\textit{double reduction} (see e.g. \cite{lynch1998} p. 57).
Double reduction happens when, as a result of recombination, replicated gene
copies on sister chromatids move to the same pole during anaphase I, as
illustrated in \cref{fig:dr}.
In the example shown in \cref{fig:dr}, one of the four generated gametes is $AA$,
which would not occur in the ordinary bivalent meiosis, because in that case,
paired chromosomes (involved in cross-overs) are separated during anaphase I.
The frequency of double reduction at a locus in the presence of multivalent
formation is hence determined by the frequency at which that locus is involved
in a cross-over (which depends on the distance to the centromere), and has an
upper bound at $1/6$ \citep{stift2008}.

When double reduction occurs, an $ABCD$ genotype would generate 10 distinct
gametes, as opposed to 6 in when chromosomes form bivalents. As a result, the
segregation variance is increased by double reduction.
For a random genotype $X_1X_2X_3X_4$, we can find the gametic segregation
variance contributed by a locus when double reduction happens as
\begin{align*}
    \Ex[\Var[Y|X_1,X_2,X_3,X_4]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3,X_4]] \\
        &= \Var[2X] - \Var\left[\frac1 4 (2X_1 + 2X_2 + 2X_3 + 2X_4)\right] \\
        &= 4v_x - \frac{1}{4}4v_x 
        = 3v_x
\end{align*}
where $X$ denotes the additive effect of a random allele at the locus drawn from the reference
population and $v_x = \Var[X]$.
In the absence of double reduction we have
\begin{align*}
    \Ex[\Var[Y|X_1,X_2,X_3,X_4]] 
        &= 2\Var[X] - \Var\left[\frac1 6 \sum_{i=1}^3 \sum_{j=i+1}^4(X_i+X_j)\right]
        = v_x
\end{align*}
Assuming that the probability of double reduction at any locus is $\alpha_4$,
and summing over independent loci, we find that the gametic segregation
variance in the presence of double reduction should be
\begin{equation}
V_{S(4,2)} = (1-\alpha_4)V_{x,4} + 3\alpha_4 V_{x,4} = V_{x,4}(1+2\alpha_4).
\label{eq:dr}
\end{equation}
where, again, $V_{x,4}$ is as defined in \cref{sec:scaling}.

\begin{figure}
\begin{center}
\begin{tikzpicture}
\footnotesize
\node (n1) at (0, 5) [] {$A\ B\ C\ D$};
\node (n2) at (0, 4) [] {$A\underline{A}\ \underline{B}\overline{B}\ CC\ \overline{D}D$};
\node (n3) at (0, 3) [] {$AB\ AD | CC\ BD$};
\node (n4) at (-2, 2) [] {$AB\ AD$};
\node (n5) at ( 2, 2) [] {$CC\ BD$};
\node (n6) at (-3, 1) [] {$AA$};
\node (n7) at (-1, 1) [] {$BD$};
\node (n8) at ( 3, 1) [] {$CB$};
\node (n9) at ( 1, 1) [] {$CD$};
\draw[->] (n1) -- (n2) node [right,midway] {\small replication};
\draw[->] (n2) -- (n3) node [right,midway] {\small recombination};
\draw[->] (n3) -- (n4);
\draw[->] (n3) -- (n5) node [right,midway] {\small \ \ meiosis I};
\draw[->] (n5) -- (n8) node [right,midway] {\small \ \ meiosis II};
\draw[->] (n5) -- (n9);
\draw[->] (n4) -- (n6);
\draw[->] (n4) -- (n7);
\end{tikzpicture}
\end{center}
\begin{caption}{
Schematic illustration of a meiotic division in an autotetraploid leading to
double reduction at a locus with genotype $ABCD$.
Two recombination events are assumed to occur at the locus (denoted by the
bars).
\label{fig:dr}
}\end{caption}
\end{figure}

\paragraph{Unreduced gamete formation in diploids.}

\begin{figure}
\begin{center}
\begin{tikzpicture}
\footnotesize
\node (rep) at (-6, 4.5) [] {\small Replication};
\node (rep) at (-6, 3.5) [] {\small Recombination};
\node (rep) at (-6, 2.5) [] {\small First division};
\node (rep) at (-6, 1.5) [] {\small Second division};
\node (rep) at (-6, 0)   [] {\small Gametes};
\node (Aa)  at (0, 5) [] {$AB$};
\node (AAaa)  at (0, 4) [] {$AA\ BB$};
\node (rAaAa) at (-2, 3) [] {$AB\ AB$};
\node (rAAaa) at ( 2, 3) [] {$AA\ BB$};
\path (Aa) edge (AAaa);
\path (AAaa) edge node[near end, above] {$c$} (rAaAa) ;
\path (AAaa) edge node[near end, above] {\qquad $1-c$} (rAAaa);

\node (fdr11) at (-3, 2) [] {$AB\ AB | \emptyset$};
\node (fdr12) at (-3, 1) [] {$A|B\  A|B$}; 
\node (AA) at (-3.8,0) [] {$AA$};
\node (p1) at (-3.8,-0.5) [] {$\frac{1}{4}$};
\node (Aa) at (-3,0) [] {$AB$};
\node (p2) at (-3,-0.5) [] {$\frac{1}{2}$};
\node (aa) at (-2.2,0) [] {$BB$};
\node (p2) at (-2.2,-0.5) [] {$\frac{1}{4}$};
\path (fdr12) edge (AA);
\path (fdr12) edge (Aa);
\path (fdr12) edge (aa);
\path (fdr11) edge (fdr12);
\path (rAaAa) edge node[left] {$f$} (fdr11);

\node (sdr11) at (-1, 2) [] {$AB | AB$}; 
\node (sdr12) at (-1, 1) [] {$AB | \emptyset$}; 
\node (Aa2) at (-1,0) [] {$AB$};
\path (sdr12) edge (Aa2);
\path (sdr11) edge (sdr12);
\path (rAaAa) edge node[right] {$1-f$} (sdr11);

\node (fdr21) at (1,2) [] {$AA\ BB|\emptyset$};
\node (fdr22) at (1,1) [] {$A|A\ B|B$};
\node (Aa3) at (1,0) [] {$AB$};
\path (rAAaa) edge node[left] {$f$} (fdr21);
\path (fdr21) edge (fdr22);
\path (fdr22) edge (Aa3);

\node (sdr21) at (3, 2) [] {$AA | BB$};
\node (sdr22) at (3, 1) [] {$AA| \emptyset, BB|\emptyset$};
\node (AA4) at (2.2,0) [] {$AA$};
\node (p1) at (2.2, -0.5) [] {$\frac{1}{2}$};
\node (aa4) at (3.8,0) [] {$BB$};
\node (p2) at (3.8, -0.5) [] {$\frac{1}{2}$};
\path (rAAaa) edge node[right] {$1-f$} (sdr21);
\path (sdr21) edge (sdr22);
\path (sdr22) edge (AA4);
\path (sdr22) edge (aa4);
\end{tikzpicture}
\begin{caption}{
Schematic representation of the different pathways for unreduced gamete
formation in diploids and their different outcomes.
\label{fig:udip}
}\end{caption}
\end{center}
\end{figure}

The mechanisms of unreduced gamete formation do not necessarily lead to a
faithful transmission of the complete diploid genome.
Unreduced gametes are formed in two ways, depending on the meiotic abberation
that leads to their origin: (1) first division restitution (FDR) of (2) second
division restitution (SDR) \citep{bretagnolle1995,storme2013}.
Consider a locus in a diploid with two distinct genes $A$ and $a$. Assume
recombination happens with probability $c$ and that conditional on unreduced
gamete formation, formation is due to FDR with probability $f$ while it is due
to SDR with probability $1-f$.
The different unreduced gametes that are formed are represented schematically
in \cref{fig:udip}.
Writing the genotype at a locus in the diploid parent as $X_1X_2$, with allelic
effects $X_1$ and $X_2$, the genotypic value of an unreduced gamete at this
locus will be
$$Y = \begin{cases}
    2X_1 & \text{w.pr.  } \alpha_2/2 \\
    2X_2 & \text{w.pr.  } \alpha_2/2 \\
    X_1 + X_2 & \text{w.pr.  } 1-\alpha_2 
    \end{cases}$$
where $\alpha_2 = 1 - f - c + \frac{3}{2}cf$ is the probability that two copies
of the same gene end up in a diploid gamete produced by a diploid individual
(see the diagram in \cref{fig:udip}). 
Conditional on the latter event, we get
\begin{align*}
\Ex[\Var[Y|X_1,X_2]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2]] \\
    &= \Var[2X] - \Var\left[\frac{1}{2}(2X_1 + 2X_2)\right] \\
    &= 2v_x
\end{align*}
Conditioning on the complementary event, all gametes have genetic value
$Y = X_1+X_2$, so that the segregation variance is 0.
Summing across independent loci, we have $V_{S(2,2)} = 2\alpha_2 V_{x,2}$.

\paragraph{Meiosis in triploids.}

Triploids, when viable, may be important for the dynamics of mixed-ploidy
populations due to the formation of a so-called triploid bridge.
The formation of triploids presents no immediate issues, we simply need to
track the segregation variance contributions from both donor gametes, and
relate these to $V_{0,3}$.
Sexual reproduction in triploids is however more complicated.
There are no known mechanisms to coordinate the assortment of chromosomes in
for instance a haploid and diploid gamete, and meiosis, if it happens, usually
results in aneuploid gametes \citep{ramsey1998}.

Experimental results indicate that, at least in yeast, triploids usually form
trivalents and undergo recombination, after which each trivalent is randomly
assorted in the daughter cells, some receiving one, others two copies of a
given chromosome \citep{charles2010}.
In the absence of gametic nonreduction, the probability
of obtaining euploid gametes (two diploid and two haploid gametes) from such 
a process is $(1/2)^n$, where $n$ is the number of chromosomes. If the number
of chromosomes is small this is not negligible, for instance in
\textit{Arabidopsis thaliana} we would have $(1/2)^5 \approx 0.03$, which is of
the same order as the unreduced gamete formation rate.
Unreduced (triploid) gametes may also be produced and important for the
dynamics of mixed-ploidy populations \citep{ramsey1998}.
However, they generate additional difficulty, since in order to compute the
contributed variance under inbreeding, we would need an additional identity
coefficient recording the probability that three genes are IBD at a locus.
We will hence ignore the possibility of unreduced gamete production in
triploids.
We note that, on the supposition that diploid gametes are produced by random
assortment of chromosomes in a haploid and diploid gametes, $\alpha_3 \le 1/4$.

When a triploid produces a haploid gamete, we get as segregation variance at a
single locus
\begin{align*}
  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
    &= v_x - \Var\left[\frac{1}{3}\left(X_1 + X_2 + X_3\right)\right] \\ 
    &= v_x - \frac{1}{9}3v_x = \frac{2}{3}v_x
\end{align*}
Summing over many loci, we have $V_{S(3,1)} = \frac{2}{3}V_{x,3}$.
When a triploid produces a diploid gamete, we assume there is, as in diploids
and tetraploids, a probability $\alpha_3$ that the same gene copy ends up twice
in the gamete.
Conditional on this happening, we have
\begin{align*}
  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
    &= \Var[2X] - \Var\left[\frac{1}{3}\left(2X_1 + 2X_2 + 2X_3\right)\right] \\
    &= 4V_x - \frac{4}{9}3V_x = \frac{8}{3}V_x
\end{align*}
Conditional on this \textit{not} happening, 
\begin{align*}
  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
    &= 2V_x - \Var\left[\frac{1}{3}\left((X_1+X_2) + (X_1+X_3) +
      (X_2+X_3)\right)\right] \\
    &= 2V_x - \frac{4}{9}3V_x = \frac{2}{3}V_x
\end{align*} 
Putting this together and summing over many loci, the segregation variance for
a diploid gamete from a triploid individual would be
  $$V_{S(3,2)} = \frac{8}{3}V_{x,3}\alpha_3 + \frac{2}{3}V_{x,3} (1-\alpha_3) 
            = \frac{2}{3}V_{x,3}(1 + 3\alpha_3)$$

The different expressions for the gametic segregation variance in the
mixed-ploidy model, adjusted for inbreeding, are summarized in
\cref{m-tbl:segvar} in the main text.
